---
title: "01 Wrangle"
output: html_document
---

Author: Kaitlin Volk; with edits by Jeff Cegan  
Date: May 19, 2021  
Project HAB with Jodi Ryder  

```{r, include=FALSE}
#load global packages and variables
source("00_global.R") #in the same folder as the .rmd file
```

### Load and filter data. 
Summary of data:

```{r, echo=FALSE}
load(paste0(dirpath,"wqdata.rda")) #so full path is ../data/wgdata.rda 
names(dat) <- c("loc_id", "sample_num", "lab_id","lab_snum", "storet_num", "units",   
                "value", "text_value", "detect","pquant", "pre_recov", "post_recov",
                "dilution", "lab_anlz", "dcl_flag", "date_entered", "revision", "val_qual", #I exported the text file without headers
                "val_id", "export_flag", "act_value","test_methd", "prep_mthd")             #so I'm adding them in manually here

target.param <- c("temp_c","DO_mgl","DO_perc","chlA_tri","chlA_spect",
                  "P_total","N_total","flow","secchi_in","TSS",           #A list of the priority parameters to be used when 
                  "TDS", "ammonia","nitrate","phosphate", "C_total")      #filtering in the next dataframe

dat2 <- dat %>%   
  mutate(param = recode(storet_num, "00010"= "temp_c",  #Replace the storet codes of the priority params with descriptive text
                                        "00299" = "DO_mgl","00301" = "DO_perc",
                                        "32210" = "chlA_tri","32211" = "chlA_spect",
                                        "00665" = "P_total","00600" = "N_total",
                                        "74073" = "flow","00077" = "secchi_in",
                                        "00515" = "TSS","00530" = "TDS",
                                        "00610" = "ammonia","00620" = "nitrate",
                                        "00650" = "phosphate","00680" = "C_total"),
         sample_num=recode(sample_num, "210808162000000.0" = "201808162000000.0"),  #fix clearly wrong date
         yr = str_sub(sample_num,1,4),  #Year column
         mo=str_sub(sample_num,5,6),    #Month column
         dy = str_sub(sample_num,7,8),  #day column
         sample_date=make_date(yr, mo, dy), #full date column
         depth = as.numeric(str_sub(sample_num,-5,-1)), #depth of sample
         reservoir = str_sub(loc_id,1,3)) %>% #reservior of sample
  select(c(1,30,2,28, 25, 26,29,24,7,6)) %>% #extract just the location, date, depth, and paramater data
  filter(param %in% target.param, !is.na(value)) #extract rows with the priority parameters only and remove NA's
summary(dat2)
```

### Exploring ChlA data with Plots
ChlA is an important variable for identifying HABs and for the correlations in 1.6 of the SOW  
Two methods have been used: trichromatic and Spectrophotomatic. Is one better? Should they both be used?

```{r, echo=FALSE}
dat.chla <- dat2 %>% filter(param == "chlA_tri"| param =="chlA_spect",!is.na(value)) #pull just the chla dat
ggplot(dat.chla, aes(x=sample_date, y=value, color=param)) + geom_line() +  #graph chla over time for the two methods
  ylim(0,200) + facet_wrap( ~ reservoir, nrow = 6)                          #facet by reservior to see differences

dat.chla.dif <- dat.chla %>%    
  pivot_wider(names_from=param, values_from=value, values_fn=mean) %>% 
  select(1,3,9,10) %>%  mutate(difference=chlA_tri-chlA_spect) %>%    #calculate difference in methods for samples with values for both methods
  filter(!is.na(difference)) %>% arrange(-difference)                   #2% of values are in disagreement

#Given the small percentage of differences, combine the two chlA columns into 1
dat3 <- dat2 %>% mutate(param = recode(param, "chlA_tri" = "chlA", "chlA_spect" = "chlA"))
dat3w <- dat3 %>% select(-10) %>% pivot_wider(names_from= param, values_from = value, values_fn=mean) 
rm(dat2) #remove previous version
```

### Look for outliers/weirdness in the variables

```{r, echo=FALSE}
ggplot(dat3w) + geom_qq(aes(sample=temp_c)) #exclude below -2 and above 40
ggplot(dat3w) + geom_qq(aes(sample=DO_mgl)) #eclude below 0 and above 50
ggplot(dat3w) + geom_qq(aes(sample=flow))   #one outlier at over 13000 cfu (next highest 5388)
ggplot(dat3w) + geom_qq(aes(sample=DO_perc)) #excle above 300
ggplot(dat3w) + geom_qq(aes(sample=TSS)) #exclude above 2000
ggplot(dat3w) + geom_qq(aes(sample=C_total)) #exclude above 100
ggplot(dat3w) + geom_qq(aes(sample=P_total)) # gonna need to mess around with this b/c of units
ggplot(dat3w) + geom_qq(aes(sample=depth)) #exclude above 200 
ggplot(dat3w) + geom_qq(aes(sample=depth)) + facet_wrap(~reservoir, nrow=6) #depth by reservoir confirm that I should exclude depths over 200
#Tested and looks passable: chla, phoshate (only 7 values), nitrate, N_total, secchi_in,TDS, ammonia
```

### Dive into total phosphorous (P_total)
```{r, echo=FALSE}
dat.p <- dat3 %>% filter(param=="P_total") %>%                               #Looking at P data only
  mutate(units=recode(units, "MG/L" = "mg/L", "mg/l" = "mg/L", "pH"="ug/L")) #simplify variations so you only have ug/L and mg/L

dat.p %>% filter(value<1) %>% ggplot(aes(x=sample_date, y=value, color=units)) + 
  geom_point() + facet_wrap(~reservoir, nrow=6)  #Look at P by reservior and units for values under 1
dat.p %>% filter(value<1000 & value >1) %>% ggplot(aes(x=sample_date, y=value, color=units)) + 
  geom_point() + facet_wrap(~reservoir, nrow=6) #Look at P by reservior and units for values over 1 but below 1000 
dat.p %>% filter(value>1000) %>% ggplot(aes(x=sample_date, y=value, color=units)) + 
  geom_point() + facet_wrap(~reservoir, nrow=6) #Look at P by reservior and for values over 1000 (the rounded upper limit in my brief lit search)
#If x is less than 1, convert to ug/l by multiplying value by 1000 
#if x > 1 but < 1000, recode mg/l to be uq/l
#if x > 1000, exclude regardless of unit
```

### Clean the data
```{r, echo=FALSE}
fine <- c("chlA", "phosphate", "nitrate","N_total", "secchi_in", "TDS", "ammonia")   #parameters that won't be cleaned

dat4 <- dat3 %>% filter((param=="temp_c" & value <40 & value> -2)|      #exclude outlying data that
                          (param=="DO_mgl" & value>0 & value <50)|      #we determined above from visualiztion
                          (param=="flow" & value <6000)|                            
                          (param=="DO_perc" & value<300)|                           
                          (param=="TSS" & value < 2000)|                             
                          (param=="C_total" & value <100)|                          
                          (param=="P_total" & value <1000)|                       
                          (param%in%fine), 
                        depth <= 200) %>% 
  mutate(value=case_when(param=="P_total" & value <1 ~ value*1000, TRUE~value), #complete proper conversion/recoding for P data
         units=case_when(param=="P_total"~ "ug/L", TRUE~as.character(units)))
rm(dat3) #remove previous version

dat4w <- dat4%>% select(-10) %>% pivot_wider(names_from= param, values_from = value, values_fn=mean)
rm(dat3w) #remove previous version
```

### Visualize the cleaned data
```{r, echo=FALSE}
ggplot(dat4w) + geom_qq(aes(sample=temp_c))
ggplot(dat4w) + geom_qq(aes(sample=DO_mgl)) 
ggplot(dat4w) + geom_qq(aes(sample=flow))   
ggplot(dat4w) + geom_qq(aes(sample=DO_perc)) 
ggplot(dat4w) + geom_qq(aes(sample=TSS)) 
ggplot(dat4w) + geom_qq(aes(sample=C_total)) 
ggplot(dat4w) + geom_qq(aes(sample=P_total))
ggplot(dat4w) + geom_qq(aes(sample=depth)) + facet_wrap(~reservoir, nrow=6)
```
### Reserviors and sites per reservoir:

```{r, echo=FALSE}
reservoir.summary <- dat4 %>% group_by(reservoir) %>% summarize(n.sites = n_distinct(loc_id))  #Get counts for the number of samples sites per reservoir
reservoir.summary 
```
Save cleaned data
```{r, echo=FALSE}
save(dat4, file = paste0(dirpath,"wqdata_cleaned.rda"))
```


